Prevalence of *Haemoproteus* parasites in American White Ibis from Southern Florida

### 2021-12-06 products/manuscript/Manuscript.Rmd
Peer Review Edits to the MS
Peer review edits from the 2 reviewers. Almost completed for submission.

## Abstract

@@

## Loading Required Data and Packages
```{r}
#Loading required packages #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(tidymodels) #to run the statistical analyses
library(ranger) #implementation of random forests
library(vip) #For plotting decision tree model
library(rpart) #For plotting machine learning models
library(rpart.plot) #For plotting machine learning models
set.seed(144)
#Loading in the complete Ibis Dataset
data_location1 <- here::here("processed_data","CompleteIbisDataset.rds")
IbisC <- readRDS(data_location1)
#Loading required materials - Figures
#Exploratory Figures
figure_Exp_1 <- here::here("results", "Figures", "ExpFigure1.png")
figure_Exp_4 <- here::here("results", "Figures", "ExpFigure4.png")
figure_Exp_6 <- here::here("results", "Figures", "ExpFigure6.png")
#Machine Learning Model Diagnostic Figures
figure_ML_4 <- here::here("results", "Figures", "ImpPlotRF.jpeg")
figure_ML_5 <- here::here("results", "Figures", "ROC_Test_RF.jpeg")
#Loading and preparing the Exploratory Analysis Tables
IbisCount1 <- IbisC %>% count(Sex, Age, wt = HaeParasitPA)
IbisCount2 <- IbisC %>% count(Season, wt = HaeParasitPA)
IbisCount3 <- IbisC %>% count(Site, wt = HaeParasitPA)
#Loading Model Results
data_locationPA4 <- here::here("results", "Tables", "Model4MetricsPA.Rda")
model4 <- readRDS(data_locationPA4)
data_locationPA5 <- here::here("results", "Tables", "Model5MetricsPA.Rda")
model5 <- readRDS(data_locationPA5)
data_locationPA8 <- here::here("results", "Tables", "Model8MetricsPA.Rda")
model8 <- readRDS(data_locationPA8)
```

## Introduction

@@

### General Background Information
The data this analysis concerns is American White Ibis (*Eudocimus albus*) blood parasite data that was collected from 2010-2017. The parasite the data focuses on is the protozoan *Haemoproteus*, which causes a chronic blood infection in White Ibises and many other birds. *Haemoproteus* is a vector-borne disease, it is transmitted by *Culicoides* biting midges when they feed on the birds and causes long-lasting parasitemia [@Coker2017]. Adult healthy birds tolerate haemosporidian infections well, but young birds and birds outside their normal range that are naive to the pathogen are more likely to develop clinical disease [@Coker2017]. Haemosporidian infections usually cause long-lasting chronic infections in natural hosts, and can cause reduced reproductive success, host fitness, and increased stress and disease susceptibility [@Coker2017]. Lower parasitemia values suggest that the infections in the Ibis are chronic [@Coker2017].

The purpose of this analysis is to determine whether the presence of *Haemoproteus* was increasing over time in Ibis in southern Florida, and if so, determine any demographic trends among the Ibis, or temporal or spatial trends among *Haemoproteus* ecology. Here we investigate whether age, sex, sampling site, capture season, date of capture, body condition, and mass of captured Ibis make significant differences on the overall presence of *Haemoproteus* in the birds.

## Methods

### Data Collection

@@

### Data Cleaning

@@

### Data Analysis

@@

### Study Population

@@

### Exploratory Analysis

@@

#### Tables
```{r}
IbisCount1
```
Overall, looking at the distribution of Ibis sex and age, the are more females than males and adults than juveniles. Female adults have the highest number with 109. Another clear observation from this table is that there are a large number of these missing variables in the dataset. This will most likely have to be completed with and without imputation depending on the model used. The results could be somewhat biased due to this distribution.

```{r}
IbisCount2
```
This table shows the number of Ibis sampled during each season. The majority of the Ibis were sampled in the later years (2015-2017) than in the earlier years (2010-2014). 13 samples did not have a recorded season. The highest number of Ibis sampled were in Spring, followed by Summer in the earlier years and Fall in the later years.

```{r}
IbisCount3
```
The majority of the Ibis were sampled in SWA, DRP, and ICP. Some sites had very low sampling numbers. Most of the Ibis were sampled in urban sites compared to natural sites.

#### Figures

@@

Figure 1: Haemoproteus parasitemia from sampled infected Ibises from 2010-2017 by habitat type
```{r}
knitr::include_graphics(figure_Exp_1, error = FALSE)
```

@@

Figure 2: Haemoproteus parasitemia by total female and male Ibis mass
```{r}
knitr::include_graphics(figure_Exp_4, error = FALSE)
```

@@

Figure 3: Haemoproteus parasitemia by sample site and habitat type
```{r}
knitr::include_graphics(figure_Exp_6, error = FALSE)
```

@@

### Statistical Analyses

#### Univariate and Multivariate Logistic Models

@@

```{r}
model4
```
This univariate logistic regression looked at the effect of sampling season on the presence of *Haemoproteus* in infected birds. The results were somewhat satisfactory with an accuracy of 0.639 and roc_auc of 0.655.

```{r}
model5
```
This univariate logistic regression looked at the effect of sampling date on the presence of *Haemoproteus* in infected birds. The results were somewhat unsatisfactory with an accuracy of 0.539 and roc_auc of 0.548.

```{r}
model8
```
This multivariate logistic regression model looked at the effect of sampling date and sampling season on the presence of Haemoproteus in infected birds. The results here had the highest performance of all the other models so far. As the sampling effort continued from 2010-2017, the passage of time and season had a significant effect on the presence of *Haemoproteus* in Ibis. However, the performance of this model isn't high enough to justify successfully concluding with that explanation.

### Random Forest Model

A random forest model was run on all the predictors against the *Haemoproteus* parasitemia presence and absence data. The number of features to consider at each split and the minimum number of data points required for the node to be split were both tuned with a tuning grid. The random forest model consisted of 500 different decision trees. The best model resulted in an roc_auc of 0.712 on the training data.

```{r Variable Importance Model}
knitr::include_graphics(figure_ML_4, error = FALSE)
```
The variable importance plot suggests that the sampling date and mass of the sampled bird in grams are the two most important predictors of *Haemoproteus* presence in the sampled Ibis.

@@

```{r ROC Curve for Random Forest Model, echo = FALSE, out.width = '100%'}
knitr::include_graphics(figure_ML_5, error = FALSE)
```

### Discussion

@@

It may also be the case that the results of the analysis suggesting sampling date as the most important predictor is simply due to the higher number of Ibis being sampled in the later years of the study. The other individual predictors were all used in univariate logistic regression models, and none of them had highly significant results.

### Conclusions

@@

>>>>

```{r, echo = FALSE, message = FALSE}
#Loading required packages #for data processing
library(here) #to set paths
library(tidyverse) #all required data manipulation packages
library(tidymodels) #to run the statistical analyses
library(ranger) #implementation of random forests
library(vip) #For plotting decision tree model
library(rpart) #For plotting machine learning models
library(rpart.plot) #For plotting machine learning models
set.seed(144)
#Loading in the complete Ibis Dataset
data_location1 <- here::here("processed_data","CompleteIbisDataset.rds")
IbisC <- readRDS(data_location1)
#Loading required materials - Figures
#Exploratory Figures
figure_Exp_1 <- here::here("results", "Figures", "ExpFigure1.png")
figure_Exp_4 <- here::here("results", "Figures", "ExpFigure4.png")
figure_Exp_6 <- here::here("results", "Figures", "ExpFigure6.png")
#Machine Learning Model Diagnostic Figures
figure_ML_4 <- here::here("results", "Figures", "ImpPlotRF.jpeg")
figure_ML_5 <- here::here("results", "Figures", "ROC_Test_RF.jpeg")
#Loading and preparing the Exploratory Analysis Tables
IbisCount1 <- IbisC %>% count(Sex, Age, wt = HaeParasitPA)
IbisCount2 <- IbisC %>% count(Season, wt = HaeParasitPA)
IbisCount3 <- IbisC %>% count(Site, wt = HaeParasitPA)
#Loading Model Results
data_locationPA4 <- here::here("results", "Tables", "Model4MetricsPA.Rda")
model4 <- readRDS(data_locationPA4)
data_locationPA5 <- here::here("results", "Tables", "Model5MetricsPA.Rda")
model5 <- readRDS(data_locationPA5)
data_locationPA8 <- here::here("results", "Tables", "Model8MetricsPA.Rda")
model8 <- readRDS(data_locationPA8)
```

# Abstract

@@

# Introduction

@@

## General Background Information
The data this analysis concerns is American White Ibis (*Eudocimus albus*) blood parasite data that was collected from 2010-2017. The parasite the data focuses on is the protozoan *Haemoproteus*, which causes a chronic blood infection in White Ibises and many other birds. *Haemoproteus* is a vector-borne disease, it is transmitted by *Culicoides* biting midges when they feed on the birds and causes long-lasting parasitemia [@Coker2017]. Adult healthy birds tolerate haemosporidian infections well, but young birds and birds outside their normal range that are naive to the pathogen are more likely to develop clinical disease [@Coker2017]. Haemosporidian infections usually cause long-lasting chronic infections in natural hosts, and can cause reduced reproductive success, host fitness, and increased stress and disease susceptibility [@Coker2017]. Lower parasitemia values suggest that the infections in the Ibis are chronic [@Coker2017]. This is a risk for the White Ibis population to be impacted, as well as multiple other host species, and these parasites could have a negative population level impact on them even if they don't cause acute mortality. Analyzing this data will help identify any demographic or spatiotemporal trends among infected Ibis in order to help possibly assist management of the species and parasite in Southern Florida.

The purpose of this analysis is to determine whether the presence of *Haemoproteus* was increasing over time in Ibis in southern Florida, and if so, determine any demographic trends among the Ibis, or temporal or spatial trends among *Haemoproteus* ecology. Here we investigate whether age, sex, sampling site, capture season, date of capture, body condition, and mass of captured Ibis make significant differences on the overall presence of *Haemoproteus* in the birds. We expect *Haemoproteus* presence to change between capture season and date, as well as body condition and overall mass of the birds.

# Methods

## Data Collection

@@

## Data Cleaning

@@

## Data Analysis

@@

## Study Population

@@

## Exploratory Analysis

@@

### Tables
```{r, echo = FALSE, message = FALSE}
IbisCount1
```
Overall, looking at the distribution of Ibis sex and age, the are more females than males and adults than juveniles. Female adults have the highest number with 109. Another clear observation from this table is that there are a large number of these missing variables in the dataset. This will most likely have to be completed with and without imputation depending on the model used. The results could be somewhat biased due to this distribution.

```{r, echo = FALSE, message = FALSE}
IbisCount2
```
This table shows the number of Ibis sampled during each season. The majority of the Ibis were sampled in the later years (2015-2017) than in the earlier years (2010-2014). 13 samples did not have a recorded season. The highest number of Ibis sampled were in Spring, followed by Summer in the earlier years and Fall in the later years.

```{r, echo = FALSE, message = FALSE}
IbisCount3
```
The majority of the Ibis were sampled in SWA, DRP, and ICP. Some sites had very low sampling numbers. Most of the Ibis were sampled in urban sites compared to natural sites.

### Figures

@@

Figure 1: Haemoproteus parasitemia from sampled infected Ibises from 2010-2017 by habitat type
```{r, echo = FALSE, message = FALSE}
knitr::include_graphics(figure_Exp_1, error = FALSE)
```

@@

Figure 2: Haemoproteus parasitemia by total female and male Ibis mass
```{r, echo = FALSE, message = FALSE}
knitr::include_graphics(figure_Exp_4, error = FALSE)
```

@@

Figure 3: Haemoproteus parasitemia by sample site and habitat type
```{r, echo = FALSE, message = FALSE}
knitr::include_graphics(figure_Exp_6, error = FALSE)
```

@@

## Statistical Analyses

### Univariate and Multivariate Logistic Models

@@

```{r, echo = FALSE, message = FALSE}
model4
```
This univariate logistic regression looked at the effect of sampling season on the presence of *Haemoproteus* in infected birds. The results were somewhat satisfactory with an accuracy of 0.639 and roc_auc of 0.655.

```{r, echo = FALSE, message = FALSE}
model5
```
This univariate logistic regression looked at the effect of sampling date on the presence of *Haemoproteus* in infected birds. The results were somewhat unsatisfactory with an accuracy of 0.539 and roc_auc of 0.548.

```{r, echo = FALSE, message = FALSE}
model8
```
This multivariate logistic regression model looked at the effect of sampling date and sampling season on the presence of Haemoproteus in infected birds. The results here had the highest performance of all the other models so far. As the sampling effort continued from 2010-2017, the passage of time and season had a significant effect on the presence of *Haemoproteus* in Ibis. However, the performance of this model isn't high enough to justify successfully concluding with that explanation.

## Random Forest Model

A random forest model was run on all the predictors against the *Haemoproteus* parasitemia presence and absence data. This model was the best of 3 machine learning models tested including a decision tree and LASSO model (refer to supplemental materials folder). The number of features to consider at each split and the minimum number of data points required for the node to be split were both tuned with a tuning grid. The random forest model consisted of 500 different decision trees. The best model resulted in an roc_auc of 0.712 on the training data.

```{r Variable Importance Model, echo = FALSE, message = FALSE}
knitr::include_graphics(figure_ML_4, error = FALSE)
```
The variable importance plot suggests that the sampling date and mass of the sampled bird in grams are the two most important predictors of *Haemoproteus* presence in the sampled Ibis.

@@

```{r ROC Curve for Random Forest Model, echo = FALSE, message = FALSE, out.width = '100%'}
knitr::include_graphics(figure_ML_5, error = FALSE)
```

## Discussion

@@

It may also be the case that the results of the analysis suggesting sampling date as the most important predictor is simply due to the higher number of Ibis being sampled in the later years of the study. The other individual predictors were all used in univariate logistic regression models, and none of them had highly significant results. These results suggest that much further research is needed, potentially continuing a multi-year study on the parasite that could provide stronger evidence for the relationaships between the variables analyzed here. Potentially looking at *Haemoproteus* presence in *Culicoides* midges in the area could be helpful as well.

## Conclusions

@@

## References
